---
title: "JID SPF — representative facial spectra by melanin-index class"
author: "Tina Lasisi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(openxlsx)
library(here)
library(patchwork)
library(KernSmooth)
library(colorSpec)
library(spacesXYZ)
library(spacesRGB)

knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.retina = 2, fig.width = 13, fig.height = 3.6, out.width = "100%"
)
OUTPUT_DIR <- here::here("output")
```

```{r load_and_compute}
# --- Load ISSA workbook (same convention as JID_SkinOptics.Rmd) ---
issa_path <- here::here("data", "ISSA_17_Jan_2025_Yan_Lu.xlsx")
issa_data <- read.xlsx(issa_path, sheet = 2)
issa <- issa_data[10:nrow(issa_data), ]
colnames(issa) <- issa_data[9, ]

body_names <- c("Back of Hand", "Cheek", "Cheek bone", "Chin", "Ear Lobe",
                "Forehead", "Inner arm", "Neck", "Nose tip", "Outer arm",
                "Palm", "Ring finger")
issa$G <- body_names[as.numeric(issa$G)]

# Reflectance, absorbance, melanin index (MI = 100 * A680)
wavelengths <- seq(400, 700, by = 10)
reflectance <- issa %>% dplyr::select(`400`:`700`) %>% mutate(across(everything(), as.numeric))
colnames(reflectance) <- paste0("R", wavelengths)
absorbance <- as.data.frame(lapply(reflectance, function(R) -log10(R / 100)))
colnames(absorbance) <- paste0("A", wavelengths)
spectrum <- cbind(issa %>% dplyr::select(A, C, G), reflectance, absorbance) %>%
  mutate(mi = 100 * A680)
colnames(spectrum)[3] <- "site"

# One representative facial spectrum per subject (Forehead preferred over Cheek)
face_spectrum <- spectrum %>%
  filter(site %in% c("Forehead", "Cheek")) %>%
  mutate(site_priority = if_else(site == "Forehead", 1, 2)) %>%
  arrange(C, site_priority, A) %>%
  group_by(C) %>% slice(1) %>% ungroup() %>%
  select(-site_priority)
```

```{r color_setup}
# --- Reflectance spectrum -> perceived skin color --------------------------
# Standard pipeline via colorSpec (spectral -> XYZ) + spacesXYZ / spacesRGB
# (XYZ -> CIELAB / sRGB), instead of hand-coded color-matching functions.
cmf_wl <- seq(400, 700, by = 10)   # our reflectance grid

# Responder = CIE D65 illuminant x CIE 1931 2-deg observer, on the 10 nm grid.
# The 2-deg observer is used because sRGB is formally defined against it, so the
# spectrum -> XYZ -> sRGB swatch path is internally consistent.
D65_eye <- colorSpec::product(colorSpec::D65.1nm, "material",
                              colorSpec::xyz1931.1nm, wavelength = cmf_wl)
# Global scale so a perfect diffuser yields Y = 100 (preserves D65 chromaticity)
xyz_scale <- 100 / as.numeric(colorSpec::product(
  colorSpec::neutralMaterial(1, wavelength = cmf_wl), D65_eye,
  wavelength = cmf_wl))[2]
# D65 reference white for CIELAB, computed for this observer
D65_white <- as.numeric(colorSpec::product(
  colorSpec::neutralMaterial(1, wavelength = cmf_wl), D65_eye,
  wavelength = cmf_wl)) * xyz_scale

# Reflectance (%) sampled on cmf_wl -> CIE XYZ (Y = 100 for a perfect diffuser)
spectrum_to_XYZ <- function(R_percent) {
  sp <- colorSpec::colorSpec(R_percent / 100, wavelength = cmf_wl,
                             quantity = "reflectance")
  as.numeric(colorSpec::product(sp, D65_eye, wavelength = cmf_wl)) * xyz_scale
}

# XYZ -> sRGB hex (clamped to gamut)
spectrum_to_hex <- function(R_percent) {
  v <- spacesRGB::RGBfromXYZ(matrix(spectrum_to_XYZ(R_percent) / 100, nrow = 1),
                             space = "sRGB")$RGB
  v <- pmin(pmax(v, 0), 1)
  grDevices::rgb(v[1], v[2], v[3])
}

# XYZ -> CIELAB (D65 reference white)
spectrum_to_lab <- function(R_percent) {
  setNames(as.numeric(spacesXYZ::LabfromXYZ(spectrum_to_XYZ(R_percent),
                                            white = D65_white)),
           c("L", "a", "b"))
}
```

```{r representative_spectra, fig.width=10, fig.height=3.5, fig.cap="Representative facial reflectance spectra chosen to span the observed melanin-index range with clear color contrast: the lightest and darkest subjects in the sample as endpoints, plus the nearest real subject to MI ≈ 45, 65, and 85 for the middle (the reflectance curve, derivative, and color swatch all come from the same individual). Top swatch = the perceived skin color for that spectrum (CIE D65 / 1931 2° observer → sRGB). Middle = full 400–700 nm reflectance curve; bottom = its first derivative dR/dλ, estimated by local polynomial regression (KernSmooth::locpoly, derivative order 1, degree 2; direct plug-in bandwidth dpill × 0.7), matching the derivative method in Pryor et al. (2026). y-axes are held constant across classes so magnitudes are comparable. Shaded band = 500–650 nm hemoglobin-feature region."}
mi_labels <- c("Low", "Intermediate low", "Intermediate", "Intermediate high", "High")
pal <- c("Low" = "#0072B2", "Intermediate low" = "#009E73", "Intermediate" = "#E69F00",
         "Intermediate high" = "#D55E00", "High" = "#7F3C8D")

# Representatives chosen to span the observed range with clear color contrast:
# the lightest and darkest real subjects in the sample as endpoints, plus the
# nearest real subject to MI ~= 45, 65, 85 for the middle (no averaging).
# Curve, derivative, and swatch below all derive from the SAME chosen subject.
mi_targets <- c(min(face_spectrum$mi), 45, 65, 85, max(face_spectrum$mi))
rep_spectrum <- purrr::map_dfr(seq_along(mi_targets), function(i) {
  face_spectrum %>%
    slice_min(abs(mi - mi_targets[i]), n = 1, with_ties = FALSE) %>%
    mutate(eumelanin_class = mi_labels[i])
}) %>%
  mutate(eumelanin_class = factor(eumelanin_class, levels = mi_labels))

# Perceived color (sRGB hex + CIELAB) computed from the FULL spectrum 400-700 nm
Rmat <- as.matrix(rep_spectrum[, paste0("R", cmf_wl)])
rep_spectrum$hex <- apply(Rmat, 1, spectrum_to_hex)
lab <- t(apply(Rmat, 1, spectrum_to_lab))
rep_spectrum$L <- lab[, "L"]; rep_spectrum$a <- lab[, "a"]; rep_spectrum$b_ <- lab[, "b"]

# Full-spectrum long form (no wavelength filtering — show the entire curve)
rep_long <- rep_spectrum %>%
  select(C, eumelanin_class, mi, matches("^R[0-9]+$")) %>%
  pivot_longer(matches("^R[0-9]+$"), names_to = "wavelength", values_to = "reflectance") %>%
  mutate(wavelength = as.numeric(str_remove(wavelength, "R"))) %>%
  arrange(eumelanin_class, wavelength)

# First derivative dR/dλ estimated by local polynomial regression, matching the
# original analysis in Pryor et al. (2026): KernSmooth::locpoly with derivative
# order 1, degree 2, and a direct plug-in bandwidth (dpill) scaled by 0.7,
# evaluated on the measured 10 nm wavelength grid (NOT a finite difference).
deriv_long <- rep_long %>%
  group_by(eumelanin_class) %>%
  group_modify(~{
    wl  <- .x$wavelength
    ref <- .x$reflectance
    bw  <- KernSmooth::dpill(wl, ref) * 0.7
    est <- KernSmooth::locpoly(x = wl, y = ref, drv = 1, degree = 2,
                               bandwidth = bw, gridsize = length(wl),
                               range.x = range(wl))
    tibble(wavelength = est$x, deriv = est$y)
  }) %>%
  ungroup()

# Shared (fixed) y-limits across all panels
refl_ylim  <- range(rep_long$reflectance, na.rm = TRUE)
deriv_ylim <- range(deriv_long$deriv, na.rm = TRUE)
classes    <- mi_labels[mi_labels %in% rep_spectrum$eumelanin_class]

build_col <- function(cls, show_y) {
  rl  <- filter(rep_long,   eumelanin_class == cls)
  dl  <- filter(deriv_long, eumelanin_class == cls)
  row <- rep_spectrum[rep_spectrum$eumelanin_class == cls, ]
  col_line <- pal[[cls]]

  swatch <- ggplot(data.frame(x = 1, y = 1)) +
    geom_tile(aes(x, y), fill = row$hex) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(title = cls, subtitle = paste0("MI ", round(row$mi))) +
    theme_void(base_size = 11) +
    theme(plot.title    = element_text(face = "bold", hjust = 0.5, size = 11),
          plot.subtitle = element_text(hjust = 0.5, size = 11, color = "grey30"),
          plot.margin   = margin(2, 2, 2, 2))

  refl <- ggplot(rl, aes(wavelength, reflectance)) +
    annotate("rect", xmin = 500, xmax = 650, ymin = -Inf, ymax = Inf,
             fill = "#fcf8ea", alpha = 0.7) +
    geom_line(color = col_line, linewidth = 1) +
    scale_x_continuous(limits = c(400, 700), breaks = seq(400, 700, 100)) +
    scale_y_continuous(limits = refl_ylim) +
    labs(x = NULL, y = if (show_y) "Reflectance (%)" else NULL) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = if (show_y) element_text() else element_blank())

  deriv <- ggplot(dl, aes(wavelength, deriv)) +
    annotate("rect", xmin = 500, xmax = 650, ymin = -Inf, ymax = Inf,
             fill = "#fcf8ea", alpha = 0.7) +
    geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
    geom_line(color = col_line, linewidth = 0.8) +
    scale_x_continuous(limits = c(400, 700), breaks = seq(400, 700, 100)) +
    scale_y_continuous(limits = deriv_ylim) +
    labs(x = "Wavelength (nm)", y = if (show_y) "dR/dλ (%/nm)" else NULL) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          axis.title.x = element_text(size = 9.5),
          axis.text.y = if (show_y) element_text() else element_blank())

  (swatch / refl / deriv) + plot_layout(heights = c(0.7, 3, 2))
}

cols <- lapply(seq_along(classes), function(i) build_col(classes[i], show_y = (i == 1)))
p <- wrap_plots(cols, nrow = 1)
p

if (dir.exists(OUTPUT_DIR)) {
  fig_base <- file.path(OUTPUT_DIR, "JID_SPF_representative_spectra")
  ggsave(paste0(fig_base, ".pdf"), p, width = 10, height = 3.5)
  ggsave(paste0(fig_base, ".png"), p, width = 10, height = 3.5, dpi = 300)
  ggsave(paste0(fig_base, ".svg"), p, width = 10, height = 3.5)
}
```

```{r color_table}
rep_spectrum %>%
  transmute(`MI class` = eumelanin_class, MI = round(mi, 1),
            `sRGB hex` = hex,
            `L*` = round(L, 1), `a*` = round(a, 1), `b*` = round(b_, 1)) %>%
  arrange(MI) %>%
  knitr::kable(caption = "Perceived skin color per representative spectrum, computed from the full 400–700 nm reflectance curve (CIE D65 illuminant, 1931 2° observer): sRGB hex for the swatch and CIELAB coordinates.")
```

## Software and session information

**Data handling.** `openxlsx` reads the ISSA reflectance workbook; `tidyverse`
(`dplyr`, `tidyr`, `stringr`, `purrr`, `ggplot2`) does the wrangling, reshaping,
and plotting; `here` resolves project-relative paths; `patchwork` composes the
per-class swatch / reflectance / derivative columns into the final figure.
`KernSmooth` (`locpoly`, `dpill`) estimates the first derivative of each
reflectance spectrum by local polynomial regression, identical to the method in
the original SkinOptics analysis (Pryor et al., 2026).

**Spectrum → color.** The perceived skin color is computed with standard
colorimetry packages:

- **`colorSpec`** — builds the responder (CIE D65 illuminant × CIE 1931 2°
  standard observer) and integrates each reflectance spectrum to CIE XYZ
  tristimulus values (`product()`), normalized so a perfect diffuser yields
  Y = 100.
- **`spacesRGB`** — converts XYZ to sRGB (`RGBfromXYZ()`) for the color swatches.
- **`spacesXYZ`** — converts XYZ to CIELAB (`LabfromXYZ()`) for the reported
  L\*, a\*, b\* coordinates.

```{r session_info}
sessionInfo()
```
